#Uma breve introdução ao desenho de pesquisa{#studydesign}

>*Consultar o estatístico após a conclusão de um experimento muitas vezes significa apenas pedir-lhe que realize um exame post-mortem. Ele talvez possa dizer do que o experimento morreu.*
>
>-- Sir Ronald Fisher^[Discurso presidencial no Primeiro Congresso Indiano de Estatística, 1938. Fonte: http://en.wikiquote.org/wiki/Ronald_Fisher]


Neste capítulo, começaremos a refletir sobre as ideias básicas que envolvem a concepção de um estudo, a coleta de dados, a verificação do funcionamento da coleta de dados e assim por diante. Não forneceremos informações suficientes para que você possa elaborar seus próprios estudos, mas forneceremos muitas das ferramentas básicas necessárias para avaliar os estudos realizados por outras pessoas. No entanto, como o foco deste livro é muito mais a análise de dados do que a coleta de dados, apresentarei apenas uma breve visão geral. Observe que este capítulo é "especial" de duas maneiras. Em primeiro lugar, é muito mais específico para a psicologia do que os capítulos posteriores. Em segundo lugar, concentra-se muito mais no problema científico da metodologia de pesquisa e muito menos no problema estatístico da análise de dados. No entanto, os dois problemas estão relacionados, por isso é tradicional que os livros didáticos de estatística discutam o problema com um pouco de detalhe. Este capítulo se baseia fortemente em @Campbell1963 e @Stevens1946 para a discussão das escalas de medida. 

##Introdução à psicometria (mensuração psicológica){#measurement}

A primeira coisa que devemos entender é que a coleta de dados pode ser considerada uma espécie de **mensuração**. Ou seja, o que estamos tentando fazer aqui é mensurar algo sobre o comportamento humano ou a mente humana. O que quero dizer com "mensuração"? 

###Algumas reflexões sobre a mensuração psicológica

A mensuração em si é um conceito sutil, mas basicamente se resume a encontrar uma maneira de atribuir números, rótulos ou algum outro tipo de descrição bem definida a "coisas". Portanto, qualquer um dos seguintes contaria como uma medição psicológica:

PAROU AQUI.
- Minha **idade** é *33* anos.
- Eu *não* **gosto de anchovas.**
- Meu **gênero cromossômico** é *masculino*.
- Meu **gênero autoidentificado** é *feminino*.???^[Well... now this is awkward, isn't it? This section is one of the oldest parts of the book, and it's outdated in a rather embarrassing way. I wrote this in 2010, at which point all of those facts *were* true. Revisiting this in 2018... well I'm not 33 any more, but that's not surprising I suppose. I can't imagine my chromosomes have changed, so I'm going to guess my karyotype was then and is now XY. The self-identified gender, on the other hand... ah. I suppose the fact that the title page now refers to me as Danielle rather than Daniel might possibly be a giveaway, but I don't typically identify as "male" on a gender questionnaire these days, and I prefer *"she/her"* pronouns as a default (it's a long story)!  I did think a little about how I was going to handle this in the book, actually. The book has a somewhat distinct authorial voice to it, and I feel like it would be a rather different work if I went back and wrote everything as Danielle and updated all the pronouns in the work. Besides, it would be a lot of work, so I've left my name as  "Dan" throughout the book, and in ant case "Dan" is a perfectly good nickname for "Danielle", don't you think? In any case, it's not a big deal. I only wanted to mention it to make life a little easier for readers who aren't sure how to refer to me. I still don't like anchovies though :-)]??? 

Na lista acima, a parte **em negrito** é "a coisa a ser medida" e a parte em itálico é "a medição em si". Na verdade, podemos sofisticar um pouco mais essa mensuração, pensando no conjunto de medições possíveis que poderiam ter surgido em cada caso:

- Minha **idade** (em anos) poderia ter sido *0, 1, 2, 3* ... etc. O limite superior para a minha idade é um pouco vago, mas na prática seria seguro dizer que a maior idade possível é *150*, já que nenhum ser humano jamais viveu tanto.
- Quando me perguntaram se eu **gosto de anchovas**, eu poderia ter dito que *gosto*, ou *não gosto*, ou *não tenho opinião*, ou *às vezes gosto.*
- Meu **gênero cromossômico** quase certamente será *masculino* ($XY$) ou *feminino* ($XX$), mas existem algumas outras possibilidades. Eu também poderia ter *síndrome de Klinfelter* ($XXY$), que é mais semelhante ao masculino do que ao feminino. E imagino que existam outras possibilidades também.
- Meu gênero **autoidentificado** também tem grande probabilidade de ser masculino ou feminino, mas não precisa ser necessariamente compatível com meu gênero cromossômico. Também posso escolher não me identificar com *nenhum* ou me chamar explicitamente de *transgênero*.

Como você pode ver, para algumas coisas (como a idade) parece bastante óbvio qual deve ser o conjunto de medidas possíveis, enquanto para outras fica um pouco complicado. Mas quero ressaltar que, mesmo no caso da idade de alguém, as coisas podem ser muito mais sutis do que isso. Por exemplo, no exemplo acima, presumi que não haveria problema em medir a idade em anos. Mas, se você for um psicólogo do desenvolvimento, isso seria muito rudimentar, e por isso costuma-se medir a idade em *anos e meses* (se uma criança tem 2 anos e 11 meses, geralmente é escrito como "2;11"). Se você estiver interessado em recém-nascidos, pode querer medir a idade em *dias desde o nascimento*, talvez até mesmo *horas desde o nascimento*. Em outras palavras, a maneira como você especifica os valores de medida permitidos faz a diferença.

Olhando mais de perto, você também pode perceber que o conceito de "idade" não é tão preciso assim. Em geral, quando dizemos "idade", implicitamente queremos dizer "o tempo decorrido desde o nascimento". Mas essa nem sempre é a maneira correta de contar o tempo. Suponha que você esteja interessado em como recém-nascidos controlam os movimentos dos olhos. Se você se interessa por crianças em período neonatal, também pode começar a se preocupar com o fato de que "nascimento" não é o único momento significativo com o qual se preocupar. Se a bebê Alice nasce 3 semanas prematura e a bebê Bianca nasce 1 semana atrasada, faria sentido dizer que elas têm a "mesma idade" se as encontrássemos "2 horas após o nascimento"? Em certo sentido, sim. Por convenção social, usamos o nascimento como referência para falar sobre idade na vida cotidiana, uma vez que define o tempo durante o qual a pessoa opera como uma entidade independente no mundo. Mas, de uma perspectiva científica, essa não é a única coisa com a qual nos importamos. Quando pensamos na biologia dos seres humanos, muitas vezes é útil pensar em nós mesmos como organismos que crescem e amadurecem desde a concepção, e dessa perspectiva, Alice e Bianca não têm a mesma idade. Portanto, você pode definir o conceito de "idade" de duas maneiras diferentes: o tempo decorrido desde a concepção e o tempo decorrido desde o nascimento. Ao lidar com adultos, não fará muita diferença, mas ao lidar com recém-nascidos, talvez faça.

Indo além dessas particularidades, há a questão da metodologia. Qual "método de medição" específico você usará para descobrir a idade de alguém? Como antes, existem muitas possibilidades diferentes:

- Você poderia simplesmente perguntar às pessoas "quantos anos você tem?". O método de autorrelato é rápido, barato e fácil. Mas só funciona com pessoas com idade suficiente para entender a pergunta, e algumas pessoas mentem sobre sua idade.
- Você poderia perguntar a uma autoridade (por exemplo, um pai/mãe) "quantos anos tem seu filho(a)?". Esse método é rápido e, ao lidar com crianças, não é tão difícil, já que o pai/mãe está quase sempre por perto. Não funciona tão bem se você quiser saber a "idade desde a concepção", já que muitos pais não conseguem dizer com certeza quando a concepção ocorreu. Para isso, você pode precisar de uma autoridade diferente (por exemplo, um obstetra).
- Você poderia consultar registros oficiais, por exemplo, certidões de nascimento ou óbito. Esse é um esforço demorado e frustrante, mas tem suas utilidades (por exemplo, se a pessoa já morreu).

### Operacionalização: definindo seu processo de medição

Todas as ideias discutidas na seção anterior estão relacionadas ao conceito de **operacionalização**. Para ser um pouco mais preciso sobre a ideia, operacionalização é o processo pelo qual pegamos um conceito significativo, mas um tanto vago, e o transformamos em uma medida precisa. O processo de operacionalização pode envolver várias coisas diferentes:

-   Ser preciso sobre o que você está tentando medir. Por exemplo, "idade" significa "tempo desde o nascimento" ou "tempo desde a concepção" no contexto da sua pesquisa?

-   Determinar qual método você usará para medi-la. Você usará autorrelato para medir a idade, perguntará a um dos pais ou consultará um registro oficial? Se estiver usando autorrelato, como formulará a pergunta?

-   Definir o conjunto de valores permitidos que a medição pode assumir. Observe que esses valores nem sempre precisam ser numéricos, embora frequentemente sejam. Ao medir a idade, os valores são numéricos, mas ainda precisamos pensar cuidadosamente sobre quais números são permitidos. Queremos a idade em anos, anos e meses, dias ou horas? Para outros tipos de medições (por exemplo, gênero), os valores não são numéricos. Mas, assim como antes, precisamos pensar sobre quais valores são permitidos. Se estamos pedindo às pessoas que informem seu gênero, quais opções permitimos que elas escolham? É suficiente permitir apenas "masculino" ou "feminino"? Você precisa de uma opção "outro"? Ou não deveríamos dar às pessoas opções específicas e, em vez disso, deixá-las responder com suas próprias palavras? E se você abrir o conjunto de valores possíveis para incluir todas as respostas verbais possíveis, como interpretaria essas respostas?
 
A operacionalização é uma tarefa complexa e não existe uma "maneira única e verdadeira" de fazê-la. A maneira como você escolhe operacionalizar o conceito informal de "idade" ou "gênero" em uma medição formal depende da finalidade da medição. Muitas vezes, você descobrirá que a comunidade de cientistas que trabalha em sua área tem algumas ideias bastante consolidadas sobre como fazer isso. Em outras palavras, a operacionalização precisa ser pensada caso a caso. No entanto, ainda que existam muitas questões específicas em cada projeto de pesquisa, existem alguns aspectos que são bastante gerais.

Antes de prosseguir, gostaria de parar um momento para esclarecer nossa terminologia e introduzir um novo termo. Aqui estão quatro coisas diferentes que estão intimamente relacionadas entre si:

-   **Um construto teórico.** É aquilo que você está tentando medir, como "idade", "gênero" ou uma "opinião". Um constructo teórico não pode ser observado diretamente e, muitas vezes, é um pouco vago.
-   **Uma medida.** Uma medida se refere ao método ou à ferramenta que você usa para realizar suas observações. Uma pergunta em uma pesquisa, uma observação comportamental ou uma tomografia cerebral podem ser considerados mensurações.
-   **Uma operacionalização.** O termo "operacionalização" se refere à conexão lógica entre a mensuração e o constructo teórico, ou ao processo pelo qual tentamos derivar uma medida de um constructo teórico.
- **Uma variável.** Finalmente, um novo termo. Uma variável é o que obtemos quando aplicamos nossa mensuração a algo no mundo. Ou seja, as variáveis são os "dados" reais que obtemos em nossos conjuntos de dados.

Na prática, até mesmo os cientistas tendem a confundir a distinção entre essas coisas, mas é muito útil tentar entender as diferenças.

## Escalas de mensuração {#scales}

Como indicado na seção anterior, o resultado de uma mensuração psicológica é chamado de variável. Mas nem todas as variáveis são do mesmo tipo, por isso é útil entender quais são os tipos existentes. Um conceito muito útil para distinguir entre diferentes tipos de variáveis é o que conhecemos como **escalas de mensuração**.

### Escala nominal

Uma variável de **escala nominal** (também chamada de variável **categórica**) é aquela em que não há uma relação particular entre as diferentes possibilidades. Para esses tipos de variáveis, não faz sentido dizer que uma delas é "maior" ou "melhor" do que qualquer outra, e absolutamente não faz sentido calcular a média entre elas. O exemplo clássico disso é a "cor dos olhos". Os olhos podem ser azuis, verdes ou castanhos, entre outras possibilidades, mas nenhuma delas é "maior" do que qualquer outra. Como resultado, seria muito estranho falar sobre uma "cor média dos olhos". Da mesma forma, o gênero também é nominal: masculino não é melhor ou pior do que feminino. Também não faz sentido tentar falar sobre um "gênero médio". Em resumo, variáveis de escala nominal são aquelas para as quais a única coisa que se pode dizer sobre as diferentes possibilidades é que elas são diferentes. Só isso.

Vamos analisar isso um pouco mais de perto. Suponha que eu estivesse pesquisando como as pessoas se deslocam para o trabalho e para casa. Uma variável que eu teria que mensurar seria o tipo de transporte que as pessoas usam para chegar ao trabalho. Essa variável "tipo de transporte" poderia ter vários valores possíveis, incluindo: "trem", "ônibus", "carro", "bicicleta". Por enquanto, vamos supor que essas quatro sejam as únicas possibilidades. Então, imagine que eu pergunte a 100 pessoas como elas chegaram ao trabalho hoje, com este resultado:
 
|Transporte|Número de pessoas|
|:-:|:-:|
| (1) Trem | 12|
| (2) Ônibus | 30|
| (3) Carro | 48|
| (4) Bicicleta | 10|


Então, qual é o meio de transporte médio? Obviamente, a resposta aqui é que não existe. É uma pergunta boba. Pode-se dizer que viajar de carro é o meio de transporte mais popular e viajar de trem é o menos popular, mas isso é tudo. Da mesma forma, observe que a ordem em que listo as opções não é muito interessante. Eu poderia ter escolhido exibir os dados como em

|Transportation|Number of people|
|:-:|:-:|
| (3) Carro | 48|
| (1) Trem | 12|
| (4) Bicicleta | 10|
| (2) Ônibus | 30|

...e nada realmente muda.

### Escala ordinal

Variáveis de **escala ordinal** têm um pouco mais de estrutura do que variáveis de escala nominal, mas não muito. Uma variável de escala ordinal é aquela em que existe uma maneira natural e significativa de ordenar as diferentes possibilidades, mas não se pode fazer nada além disso. O exemplo usual de uma variável ordinal é "posição final em uma corrida". Você *pode* dizer que a pessoa que terminou em primeiro foi mais rápida do que a pessoa que terminou em segundo, mas *não* sabe quanto mais rápido. Como consequência, sabemos que o 1$^\circ$ > 2$^\circ$, e sabemos que o 2$^\circ$ > 3$^\circ$, mas a diferença entre o 1$^\circ$ e o 2$^\circ$ pode ser muito maior do que a diferença entre o 2$^\circ$ e o 3$^\circ$.

Aqui está um exemplo psicologicamente mais interessante. Suponha que eu esteja interessado nas atitudes das pessoas em relação às mudanças climáticas. Então, peço a algumas pessoas que escolham a afirmação (dentre quatro afirmações listadas) que mais se aproxima de suas crenças:

>(1) As temperaturas estão subindo devido à atividade humana
>(2) As temperaturas estão subindo, mas não sabemos por quê
>(3) As temperaturas estão subindo, mas não por causa dos humanos
>(4) As temperaturas não estão subindo

Observe que essas quatro afirmações, na verdade, têm uma ordem natural, em termos de "até que ponto concordam com a ciência atual". A afirmação 1 é uma correspondência próxima, a afirmação 2 é uma correspondência razoável, a afirmação 3 não é uma correspondência muito boa e a afirmação 4 está em forte oposição à ciência atual. Portanto, em termos do que me interessa (até que ponto as pessoas endossam a ciência), posso ordenar os itens como 1 > 2 > 3 > 4. Como essa ordem existe, seria muito estranho listar as opções assim...

>(3) As temperaturas estão subindo, mas não por causa dos humanos
>(1) As temperaturas estão subindo devido à atividade humana
>(4) As temperaturas não estão subindo
>(2) As temperaturas estão subindo, mas não sabemos por quê

... porque parece violar a "estrutura" natural da pergunta.

Então, vamos supor que eu fizesse essas perguntas a 100 pessoas e obtivesse as respostas abaixo:

|Resposta                                               | Núemro|
|--------                                               |:-----:|
|(1) As temperaturas estão subindo devido à atividade humana | 51    |
|(2) As temperaturas estão subindo, mas não sabemos por quê  | 20    |
|(3) As temperaturas estão subindo, mas não por causa dos humanos | 10    |
|(4) As temperaturas não estão subindo                        | 19    |

Ao analisar esses dados, parece bastante razoável tentar agrupar (1), (2) e (3) e dizer que 81 em cada 100 pessoas estavam dispostas a endossar, pelo menos parcialmente, a ciência. E também é bastante razoável agrupar (2), (3) e (4) e dizer que 49 em cada 100 pessoas registraram *pelo menos alguma discordância* com a visão científica dominante. E seria perfeitamente razoável tentar agrupar (1), (2) e (4) e dizer que 90 em cada 100 pessoas disseram... o quê? Não há nada razoável que permita agrupar essas respostas.

Dito isso, observe que, embora *possamos* usar a ordem natural desses itens para construir agrupamentos razoáveis, o que *não podemos* fazer é calcular a média deles. Por exemplo, no meu exemplo simples aqui, a resposta "média" para a pergunta é 1,97. Se você puder me dizer o que isso significa, eu adoraria saber, porque me parece algo sem lógica!

### Escala intervalar

Ao contrário das variáveis de escala nominal e ordinal, as variáveis de **escala intervalar** e de escala de razão são variáveis para as quais o valor numérico é genuinamente significativo. No caso de variáveis de escala intervalar, as *diferenças* entre os números são interpretáveis, mas a variável não tem um valor zero "natural". Um bom exemplo de uma variável de escala intervalar é a medição da temperatura em graus Celsius. Por exemplo, se ontem foram 15$^{\circ}$ e hoje foram 18$^{\circ}$, então a diferença de 3$^{\circ}$ entre as duas medições é genuinamente significativa. Além disso, essa diferença de 3$^{\circ}$ é *exatamente a mesma* que a diferença de 3$^{\circ}$ entre 7$^{\circ}$ e 10$^{\circ}$. Em resumo, adição e subtração são significativas para variáveis de escala intervalar.???^[Actually, I've been informed by readers with greater physics knowledge than I that temperature isn't strictly an interval scale, in the sense that the amount of energy required to heat something up by 3$^\circ$ depends on it's current temperature. So in the sense that physicists care about, temperature isn't actually interval scale. But it still makes a cute example, so I'm going to ignore this little inconvenient truth.]??? 

No entanto, observe que 0$^{\circ}$ não significa "nenhuma temperatura". Na verdade, significa "a temperatura na qual a água congela", o que é bastante arbitrário. Consequentemente, torna-se inútil tentar multiplicar e dividir temperaturas. É errado dizer que 20$^{\circ}$ é duas vezes mais quente que 10$^{\circ}$, assim como é estranho e sem sentido tentar afirmar que 20$^{\circ}$ é menos duas vezes mais quente que -10$^{\circ}$.

Novamente, vejamos um exemplo mais psicológico. Suponha que eu esteja interessado em observar como as atitudes de estudantes universitários do primeiro ano mudaram ao longo do tempo. Obviamente, vou querer registrar o ano em que cada aluno começou. Esta é uma variável de escala intervalar. Um aluno que começou em 2003 chegou 5 anos antes de um aluno que começou em 2008. No entanto, seria completamente insensato da minha parte dividir 2008 por 2003 e dizer que o segundo aluno começou "1,0024 vezes depois" do primeiro. Isso não faz o menor sentido.

### Escala de proporção

O quarto e último tipo de variável a ser considerado é uma variável de **escala de razão**, na qual zero significa zero, e não há problema em multiplicar e dividir. Um bom exemplo psicológico de uma variável de escala de razão é o tempo de resposta (TR). Em muitas tarefas, é muito comum registrar o tempo que alguém leva para resolver um problema ou responder a uma pergunta, porque isso é um indicador de quão difícil a tarefa é. Suponha que Alan leve 2,3 segundos para responder a uma pergunta, enquanto Ben leva 3,1 segundos. Assim como em uma variável de escala intervalar, adição e subtração são significativas aqui. Ben realmente levou 3,1 - 2,3 = 0,8 segundos a mais do que Alan. No entanto, observe que multiplicação e divisão também fazem sentido aqui: Ben levou 3,1/2,3 = 1,35 vezes mais tempo que Alan para responder à pergunta. E o motivo pelo qual você pode fazer isso é que, para uma variável de uma escala de razão como TR, "zero segundos" realmente significa "tempo nenhum".

### Variáveis contínuas versus discretas

Há um segundo tipo de distinção que você precisa conhecer, em relação aos tipos de variáveis que você pode encontrar. É a distinção entre variáveis contínuas e variáveis discretas (@tbl-tab2-4). A diferença entre elas é a seguinte:

- Uma **variável contínua** é aquela em que, para quaisquer dois valores que você possa imaginar, é sempre logicamente possível ter outro valor entre eles.
- Uma **variável discreta** é, na verdade, uma variável que não é contínua. Para uma variável discreta, é possível que não haja nada no meio.

Essas definições provavelmente parecem um pouco abstratas, mas são bem simples quando você vê alguns exemplos. Por exemplo, o tempo de resposta é contínuo. Se Alan leva 3,1 segundos e Ben leva 2,3 segundos para responder a uma pergunta, então o tempo de resposta de Cameron estará entre os dois se ele levou 3,0 segundos. E, claro, também seria possível que David levasse 3,031 segundos para responder, o que significa que seu RT estaria entre o de Cameron e o de Alan. E embora na prática seja impossível medir o RT com tanta precisão, em princípio, isso é possível. Como sempre podemos encontrar um novo valor para o RT entre quaisquer dois outros, consideramos o RT uma medida contínua.

Variáveis discretas ocorrem quando essa regra é violada. Por exemplo, variáveis de escala nominal são sempre discretas. Não existe um tipo de transporte que se situe "entre" trens e bicicletas, não da forma matemática estrita como 2,3 se situa entre 2 e 3. Portanto, o tipo de transporte é discreto. Da mesma forma, variáveis de escala ordinal são sempre discretas. Embora "2$^\circ$ lugar" se situe entre "1$^\circ$ lugar" e "3$^\circ$ lugar", não há nada que possa logicamente se situar entre "1º lugar" e "2º lugar". Variáveis de escala intervalar e de escala de razão podem ir em qualquer direção. Como vimos acima, o tempo de resposta (uma variável de escala de razão) é contínuo. A temperatura em graus Celsius (uma variável de escala intervalar) também é contínua. No entanto, o ano em que você estudou (uma variável de escala intervalar) é discreto. Não há ano entre 2002 e 2003. O número de questões que você acerta em um teste de verdadeiro ou falso (uma variável de escala de razão) também é discreto. Como uma questão de verdadeiro ou falso não permite que você esteja "parcialmente correto", não há nada entre 5/10 e 6/10. A tabéla \@ref(tab:scalescont) resume a relação entre as escalas de medida e a distinção discreta/continua. As células marcadas correspondem a coisas que são possíveis. Estou tentando enfatizar esse ponto porque (a) alguns livros didáticos erram nisso e (b) as pessoas frequentemente dizem coisas como "variável discreta" quando querem dizer "variável de escala nominal". É uma pena.

```{r scalescont, tidy=FALSE, echo=FALSE}
knitr::kable(
  rbind(c("", "contínua", "discreta"),
        c("nominal", "", "$\\checkmark$"),
        c("ordinal", "", "$\\checkmark$"),
        c("intervalar", "$\\checkmark$", "$\\checkmark$"),
        c("proporção", "$\\checkmark$", "$\\checkmark$")
    
  )
  , caption = 'A relação entre escalas de medidas e a distinção discreta/contínua. As células com marca de ???tiquecheck??? correspondem à coisas possíveis.', align="lcc",
  booktabs = TRUE
)

```

### Algumas ???complexidades???

Ok, eu sei que você vai ficar chocado ao ouvir isso, mas o mundo real é muito mais confuso do que este pequeno esquema de classificação sugere. Muito poucas variáveis na vida real realmente se enquadram nessas categorias bem definidas, portanto, você precisa ser cuidadoso para não tratar as escalas de medição como se fossem regras imutáveis. Não funciona assim. Elas são diretrizes, destinadas a ajudá-lo a pensar sobre as situações em que você deve tratar variáveis diferentes de forma diferente. Nada além disso.

Então, vamos pegar um exemplo clássico, talvez *o* exemplo clássico de uma ferramenta de medição psicológica: a **escala Likert**. A humilde escala Likert é a ferramenta básica de todo o desenho de pesquisa. Você mesmo já preencheu centenas (talvez milhares) de pesquisas que utilizam essa escala, e talvez já a tenha usado em uma pesquisa sua. Suponha que tenhamos uma pergunta numa pesquisa mais ou menos assim:

>Qual das seguintes opções melhor descreve sua opinião sobre a afirmação de que "todos os piratas são absolutamente incríveis" ...

e então as opções apresentadas ao participante são estas:

>(1) Discordo totalmente
>(2) Discordo
>(3) Nem concordo nem discordo
>(4) Concordo
>(5) Concordo totalmente

Este conjunto de itens é um exemplo de uma escala Likert de 5 pontos, na qual as pessoas são solicitadas a escolher entre uma das várias (neste caso, 5) possibilidades claramente ordenadas, geralmente com um descritor verbal para cada opção. No entanto, não é necessário que todos os itens sejam explicitamente descritos. Este é um exemplo perfeitamente bom de uma escala Likert de 5 pontos também:

>(1) Discordo totalmente
>(2) 
>(3) 
>(4) 
>(5) Concordo totalmente

As escalas Likert são ferramentas muito úteis, embora um tanto limitadas. A questão é que tipo de variável elas são? Elas são obviamente discretas, já que você não pode dar uma resposta de 2,5. Elas obviamente não são de escala nominal, já que os itens são ordenados; e também não são de escala de razão, já que não há um zero natural.

Mas elas são escalas do tipo ordinal ou do tipo intervalar? Por um lado, não podemos realmente provar que a diferença entre "concordo totalmente" e "concordo" tem o mesmo tamanho que a diferença entre "concordo" e "nem concordo nem discordo". Na verdade, na vida cotidiana é bastante óbvio que não são a mesma coisa. Isso sugere que devemos tratar as escalas Likert como variáveis ordinais. Por outro lado, na prática, a maioria dos participantes parece levar a sério a parte de "em uma escala de 1 a 5", e eles tendem a agir como se as diferenças entre as cinco opções de resposta fossem bastante semelhantes entre si. Como consequência, muitos pesquisadores tratam os dados da escala Likert como escala de intervalo.[^02-a-brief-introduction-to-research-design-3] Não é uma escala de intervalo, mas na prática é perto o suficiente para que geralmente pensemos nela como sendo uma **escala quase-intervalar**.

## Avaliando a confiabilidade de uma medição {#reliability}

Neste ponto, pensamos um pouco sobre como operacionalizar um construto teórico e, assim, criar uma medida psicológica. E vimos que, aplicando mensurações psicológicas, geramos variáveis que podem ser de muitos tipos diferentes. Neste ponto, devemos começar a discutir a pergunta óbvia: a medição é boa? Faremos isso em termos de duas ideias relacionadas: *confiabilidade* e *validade*. Em resumo, a **_confiabilidade_** de uma medição diz o quão consistentes são os resultados da medição, enquanto a **validade** de uma medida diz o quão acurada (precisa) a medida é. Nesta seção, falaremos sobre confiabilidade; falaremos sobre validade na seção ???[Avaliando a validade de um estudo]???.

A confiabilidade é, na verdade, um conceito muito simples. Refere-se à repetibilidade ou consistência de sua medição. A medição do meu peso por meio de uma "balança de banheiro" é muito confiável. Se eu subir e descer da balança várias vezes, ela continuará me dando a mesma resposta. Medir minha inteligência por meio de "perguntar à minha mãe" é muito pouco confiável. Alguns dias ela me diz que sou um pouco burro, e em outros dias ela me diz que sou um completo idiota. Observe que este conceito de confiabilidade é diferente da questão de saber se as medições estão corretas (a correção de uma medição se relaciona com sua validade). Se eu estiver segurando um saco de batatas quando subir e descer da balança, a medição ainda será confiável: ela sempre me dará a mesma resposta. No entanto, esta resposta altamente confiável não corresponde ao meu peso real, portanto, está errada. Em termos técnicos, esta é uma medição **confiável, mas inválida**. Da mesma forma, embora a estimativa da minha mãe sobre minha inteligência seja pouco confiável (consistente), ela pode estar certa. Talvez eu não seja muito inteligente, e então, apesar de a estimativa dela sobre minha inteligência flutuar bastante de um dia para o outro, em geral ela parece estar certa. Isso seria uma medida **não confiável, mas válida**. É claro que, se as estimativas da minha mãe forem muito pouco confiáveis, será muito difícil descobrir qual de suas muitas afirmações sobre minha inteligência é realmente a correta. Até certo ponto, uma medida muito pouco confiável tende a acabar sendo inválida para fins práticos; tanto que muitas pessoas diriam que a confiabilidade é necessária (mas não suficiente) para garantir a validade.

Ok, agora que sabemos claramente qual a diferença entre confiabilidade e validade, vamos pensar sobre as diferentes maneiras pelas quais poderíamos medidr a confiabilidade:

- **Confiabilidade teste-reteste**. Isso se relaciona com a consistência ao longo do tempo. Se repetirmos a medição em uma data posterior, obtemos a mesma resposta?
- **Confiabilidade entre avaliadores**. Isso se relaciona com a consistência entre as pessoas. Se outra pessoa repetir a medição (por exemplo, outra pessoa avaliar minha inteligência), ela produzirá a mesma resposta?
- **Confiabilidade de formas paralelas**. Isso se relaciona com a consistência entre medições teoricamente equivalentes. Se eu usar um conjunto diferente de balanças de banheiro para medir meu peso, ele dá a mesma resposta?
- **Confiabilidade de consistência interna**. Se uma medição é construída a partir de muitas partes diferentes que desempenham funções semelhantes (por exemplo, o resultado de um questionário sobre personalidade é obtido através do somatório de várias perguntas), as partes individuais tendem a dar respostas semelhantes (no nosso exemplo, as várias perguntas devem ter valores semelhantes). Analisaremos essa forma particular de confiabilidade mais tarde no livro, em @sec-Internal-consistency-reliability-analysis.
 
Nem todas as medições precisam possuir todas as formas de confiabilidade. Por exemplo, a avaliação educacional pode ser pensada como uma forma de medição. Uma das disciplinas que eu leciono, *Ciência Cognitiva Computacional*, tem uma estrutura de avaliação que tem um componente de pesquisa e um componente de avaliação (além de outras coisas). O componente de avaliação é *destinado* a medir algo diferente do componente de pesquisa, portanto, a medição como um todo tem baixa consistência interna. No entanto, dentro da avaliação existem várias perguntas que se destinam a (aproximadamente) medir as mesmas coisas, e essas tendem a produzir resultados semelhantes. Portanto, a avaliação por si só tem uma consistência interna bastante alta. O que é como deveria ser. Você só deve exigir confiabilidade naquelas situações em que deseja medir a mesma coisa!

##The "role" of variables: predictors and outcomes{#ivdv}

Okay, I've got one last piece of terminology that I need to explain to you before moving away from variables. Normally, when we do some research we end up with lots of different variables. Then, when we analyse our data we usually try to explain some of the variables in terms of some of the other variables. It's important to keep the two roles "thing doing the explaining" and "thing being explained" distinct. So let's be clear about this now. Firstly, we might as well get used to the idea of using mathematical symbols to describe variables, since it's going to happen over and over again. Let's denote the "to be explained" variable $Y$, and denote the variables "doing the explaining" as $X_1$, $X_2$, etc. 

Now, when we doing an analysis, we have different names for $X$ and $Y$, since they play different roles in the analysis. The classical names for these roles are **_independent variable_** (IV) and **_dependent variable_** (DV). The IV is the variable that you use to do the explaining (i.e., $X$) and the DV is the variable being explained (i.e., $Y$). The logic behind these names goes like this: if there really is a relationship between $X$ and $Y$ then we can say that $Y$ depends on $X$, and if we have designed our study "properly" then $X$ isn't dependent on anything else. However, I personally find those names horrible: they're hard to remember and they're highly misleading, because (a) the IV is never actually "independent of everything else" and (b) if there's no relationship, then the DV doesn't actually depend on the IV. And in fact, because I'm not the only person who thinks that IV and DV are just awful names, there are a number of alternatives that I find more appealing. The terms that I'll use in these notes are ***predictors*** and ***outcomes***. The idea here is that what you're trying to do is use $X$ (the predictors) to make guesses about $Y$ (the outcomes).^[Annoyingly, though, there's a lot of different names used out there. I won't list all of them -- there would be no point in doing that -- other than to note that R often uses "response variable" where I've used "outcome", and a traditionalist would use "dependent variable". Sigh. This sort of terminological confusion is very common, I'm afraid.] This is summarised in Table \@ref(tab:ivdv).


```{r ivdv, tidy=FALSE, echo=FALSE}
knitr::kable(rbind
             (c("to be explained", "dependent variable (DV)", "outcome"),
             c("to do the explaining", "independent variable (IV)", "predictor")),
caption = 'The terminology used to distinguish between different roles that a variable can play when analysing a data set. Note that this book will tend to avoid the classical terminology in favour of the newer names.',col.names = c(
  "role of the variable", "classical name","modern name"),
  booktabs = TRUE
)
```

##Experimental and non-experimental research{#researchdesigns}

One of the big distinctions that you should be aware of is the distinction between "experimental research" and "non-experimental research". When we make this distinction, what we're really talking about is the degree of control that the researcher exercises over the people and events in the study.

###Experimental research

The key features of **_experimental research_** is that the researcher controls all aspects of the study, especially what participants experience during the study. In particular, the researcher manipulates or varies the predictor variables (IVs), and then allows the outcome variable (DV) to vary naturally. The idea here is to deliberately vary the predictors (IVs) to see if they have any causal effects on the outcomes. Moreover, in order to ensure that there's no chance that something other than the predictor variables is causing the outcomes, everything else is kept constant or is in some other way "balanced" to ensure that they have no effect on the results. In practice, it's almost impossible to *think* of everything else that might have an influence on the outcome of an experiment, much less keep it constant. The standard solution to this is ***randomisation***: that is, we randomly assign people to different groups, and then give each group a different treatment (i.e., assign them different values of the predictor variables). We'll talk more about randomisation later in this course, but for now, it's enough to say that what randomisation does is minimise (but not eliminate) the chances that there are any systematic difference between groups. 

Let's consider a very simple, completely unrealistic and grossly unethical example. Suppose you wanted to find out if smoking causes lung cancer. One way to do this would be to find people who smoke and people who don't smoke, and look to see if smokers have a higher rate of lung cancer. This is *not* a proper experiment, since the researcher doesn't have a lot of control over who is and isn't a smoker. And this really matters: for instance, it might be that people who choose to smoke cigarettes also tend to have poor diets, or maybe they tend to work in asbestos mines, or whatever. The point here is that the groups (smokers and non-smokers) actually differ on lots of things, not *just* smoking. So it might be that the higher incidence of lung cancer among smokers is caused by something else, not by smoking per se. In technical terms, these other things (e.g. diet) are called "confounds", and we'll talk about those in just a moment. 

In the meantime, let's now consider what a proper experiment might look like. Recall that our concern was that smokers and non-smokers might differ in lots of ways. The solution, as long as you have no ethics, is to *control* who smokes and who doesn't. Specifically, if we randomly divide participants into two groups, and force half of them to become smokers, then it's very unlikely that the groups will differ in any respect other than the fact that half of them smoke. That way, if our smoking group gets cancer at a higher rate than the non-smoking group, then we can feel pretty confident that (a) smoking does cause cancer and (b) we're murderers. 

###Non-experimental research

**_Non-experimental research_** is a broad term that covers "any study in which the researcher doesn't have quite as much control as they do in an experiment". Obviously, control is something that scientists like to have, but as the previous example illustrates, there are lots of situations in which you can't or shouldn't try to obtain that control. Since it's  grossly unethical (and almost certainly criminal) to force people to smoke in order to find out if they get cancer, this is a good example of a situation in which you really shouldn't try to obtain experimental control. But there are other reasons too. Even leaving aside the ethical issues, our "smoking experiment" does have a few other issues. For instance, when I suggested that we "force" half of the people to become smokers, I must have been talking about *starting* with a sample of non-smokers, and then forcing them to become smokers. While this sounds like the kind of solid, evil experimental design that a mad scientist would love, it might not be a very sound way of investigating the effect in the real world. For instance, suppose that smoking only causes lung cancer when people have poor diets, and suppose also that people who normally smoke do tend to have poor diets. However, since the "smokers" in our experiment aren't "natural" smokers (i.e., we forced non-smokers to become smokers; they didn't take on all of the other normal, real life characteristics that smokers might tend to possess) they probably have better diets. As such, in this silly example they wouldn't get lung cancer, and our experiment will fail, because it violates the structure of the "natural" world (the technical name for this is an "artifactual" result; see later).

One distinction worth making between two types of non-experimental research is the difference between ***quasi-experimental research*** and ***case studies***. The example I discussed earlier -- in which we wanted to examine incidence of lung cancer among smokers and non-smokers, without trying to control who smokes and who doesn't -- is a quasi-experimental design. That is, it's the same as an experiment, but we don't control the predictors (IVs). We can still use statistics to analyse the results, it's just that we have to be a lot more careful.

The alternative approach, case studies, aims to provide a very detailed description of one or a few instances. In general, you can't use statistics to analyse the results of case studies, and it's usually very hard to draw any general conclusions about "people in general" from a few isolated examples. However, case studies are very useful in some situations. Firstly, there are situations where you don't have any alternative: neuropsychology has this issue a lot. Sometimes, you just can't find a lot of people with brain damage in a specific area, so the only thing you can do is describe those cases that you do have in as much detail and with as much care as you can. However, there's also some genuine advantages to case studies: because you don't have as many people to study, you have the ability to invest lots of time and effort trying to understand the specific factors at play in each case. This is a very valuable thing to do. As a consequence, case studies can complement the more statistically-oriented approaches that you see in experimental and quasi-experimental designs. We won't talk much about case studies in these lectures, but they are nevertheless very valuable tools!


##Assessing the validity of a study{#validity}

More than any other thing, a scientist wants their research to be "valid". The conceptual idea behind ***validity*** is very simple: can you trust the results of your study? If not, the study is invalid. However, while it's easy to state, in practice it's much harder to check validity than it is to check reliability. And in all honesty, there's no precise, clearly agreed upon notion of what validity actually is. In fact, there's lots of different kinds of validity, each of which raises it's own issues, and not all forms of validity are relevant to all studies. I'm going to talk about five different types:

- Internal validity
- External validity
- Construct validity
- Face validity
- Ecological validity

To give you a quick guide as to what matters here... (1) Internal and external validity are the most important, since they tie directly to the fundamental question of whether your study really works. (2) Construct validity asks whether you're measuring what you think you are. (3) Face validity isn't terribly important except insofar as you care about "appearances". (4) Ecological validity is a special case of face validity that corresponds to a kind of appearance that you might care about a lot.

###Internal validity

**_Internal validity_** refers to the extent to which you are able draw the correct conclusions about the causal relationships between variables. It's called "internal" because it refers to the relationships between things "inside" the study.  Let's illustrate the concept with a simple example. Suppose you're interested in finding out whether a university education makes you write better. To do so, you get a group of first year students, ask them to write a 1000 word essay, and count the number of spelling and grammatical errors they make. Then you find some third-year students, who obviously have had more of a university education than the first-years, and repeat the exercise. And let's suppose it turns out that the third-year students produce fewer errors. And so you conclude that a university education improves writing skills. Right? Except... the big problem that you have with this experiment is that the third-year students are older, and they've had more experience with writing things. So it's hard to know for sure what the causal relationship is: Do older people write better? Or people who have had more writing experience? Or people who have had more education? Which of the above is the true *cause* of the superior performance of the third-years? Age? Experience? Education?  You can't tell.  This is an example of a failure of internal validity, because your study doesn't properly tease apart the *causal* relationships between the different variables. 



###External validity

**_External validity_** relates to the **_generalisability_** of your findings. That is, to what extent do you expect to see the same pattern of results in "real life" as you saw in your study. To put it a bit more precisely, any study that you do in psychology will involve a fairly specific set of questions or tasks, will occur in a specific environment, and will involve participants that are drawn from a particular subgroup. So, if it turns out that the results don't actually generalise to people and situations beyond the ones that you studied, then what you've got is a lack of external validity.

The classic example of this issue is the fact that a very large proportion of studies in psychology will use undergraduate psychology students as the participants. Obviously, however, the researchers don't care *only* about psychology students; they care about people in general. Given that, a study that uses only psych students as participants always carries a risk of lacking external validity. That is, if there's something "special" about psychology students that makes them different to the general populace in some *relevant* respect, then we may start worrying about a lack of external validity.

That said, it is absolutely critical to realise that a study that uses only psychology students does not necessarily have a problem with external validity. I'll talk about this again later, but it's such a common mistake that I'm going to mention it here. The external validity is threatened by the choice of population if (a) the population from which you sample  your participants is very narrow (e.g., psych students), and (b) the narrow population that you sampled from is systematically different from the general population, *in some respect that is relevant to the psychological phenomenon that you intend to study*. The italicised part is the bit that lots of people forget: it is true that psychology undergraduates differ from the general population in lots of ways, and so a study that uses only psych students *may* have problems with external validity. However, if those differences aren't very relevant to the phenomenon that you're studying, then there's nothing to worry about. To make this a bit more concrete, here's two extreme examples:

- You want to measure "attitudes of the general public towards psychotherapy", but all of your participants are psychology students. This study would almost certainly have a problem with external validity.
- You want to measure the effectiveness of a visual illusion, and your participants are all psychology students. This study is very unlikely to have a problem with external validity


Having just spent the last couple of paragraphs focusing on the choice of participants (since that's the big issue that everyone tends to worry most about), it's worth remembering that external validity is a broader concept. The following are also examples of things that might pose a threat to external validity, depending on what kind of study you're doing:

- People might answer a "psychology questionnaire" in a manner that doesn't reflect what they would do in real life.
- Your lab experiment on (say) "human learning" has a different structure to the learning problems people face in real life.



###Construct validity

**_Construct validity_** is basically a question of whether you're measuring what you want to be measuring.  A measurement has good construct validity if it is actually measuring the correct theoretical construct, and bad construct validity if it doesn't.  To give very simple (if ridiculous) example, suppose I'm trying to investigate the rates with which university students cheat on their exams. And the way I attempt to measure it is by asking the cheating students to stand up in the lecture theatre so that I can count them. When I do this with a class of 300 students, 0 people claim to be cheaters. So I therefore conclude that the proportion of cheaters in my class is 0%. Clearly this is a bit ridiculous. But the point here is not that this is a very deep methodological example, but rather to explain what construct validity is. The problem with my measure is that while I'm *trying* to measure "the proportion of people who cheat" what I'm actually measuring is "the proportion of people stupid enough to own up to cheating, or bloody minded enough to pretend that they do". Obviously, these aren't the same thing! So my study has gone wrong, because my measurement has very poor construct validity.



###Face validity

**_Face validity_** simply refers to whether or not a measure "looks like" it's doing what it's supposed to, nothing more. If I design a test of intelligence, and people look at it and they say "no, that test doesn't measure intelligence", then the measure lacks face validity. It's as simple as that. Obviously, face validity isn't very important from a pure scientific perspective. After all, what we care about is whether or not the measure *actually* does what it's supposed to do, not whether it *looks like* it does what it's supposed to do. As a consequence, we generally don't care very much about face validity. That said, the concept of face validity serves three useful pragmatic  purposes:

- Sometimes, an experienced scientist will have a "hunch" that a particular measure won't work. While these sorts of hunches have no strict evidentiary value, it's often worth paying attention to them. Because often times people have knowledge that they can't quite verbalise, so there might be something to worry about even if you can't quite say why. In other words, when someone you trust criticises the face validity of your study, it's worth taking the time to think more carefully about your design to see if you can think of reasons why it might go awry. Mind you, if you don't find any reason for concern, then you should probably not worry: after all, face validity really doesn't matter much.
- Often (very often), completely uninformed people will also have a "hunch" that your research is crap. And they'll criticise it on the internet or something. On close inspection, you'll often notice that these criticisms are actually focused entirely on how the study "looks", but not on anything deeper. The concept of face validity is useful for gently explaining to people that they need to substantiate their arguments further. 
- Expanding on the last point, if the beliefs of untrained people are critical (e.g., this is often the case for applied research where you actually want to convince policy makers of something or other) then you *have* to care about face validity. Simply because -- whether you like it or not -- a lot of people will use face validity as a proxy for real validity. If you want the government to change a law on scientific, psychological grounds, then it won't matter how good your studies "really" are. If they lack face validity, you'll find that politicians ignore you. Of course, it's somewhat unfair that policy often depends more on appearance than fact, but that's how things go.
 



###Ecological validity

**_Ecological validity_** is a different notion of validity, which is similar to external validity, but less important. The idea is that, in order to be ecologically valid, the entire set up of the study should closely approximate the real world scenario that is being investigated. In a sense, ecological validity is a kind of face validity -- it relates mostly to whether the study "looks" right, but with a bit more rigour to it. To be ecologically valid, the study has to look right in a fairly specific way. The idea behind it is the intuition that a study that is ecologically valid is more likely to be externally valid. It's no guarantee, of course. But the nice thing about ecological validity is that it's much easier to check whether a study is ecologically valid than it is to check whether a study is externally valid. An simple example would be eyewitness identification studies. Most of these studies tend to be done in a university setting, often with fairly simple array of faces to look at rather than a line up. The length of time between seeing the "criminal" and being asked to identify the suspect in the "line up" is usually shorter. The "crime" isn't real, so there's no chance that the witness being scared, and there's no police officers present, so there's not as much chance of feeling pressured. These things all mean that the study *definitely* lacks ecological validity. They might (but might not) mean that it also lacks external validity.



##Confounds, artifacts and other threats to validity

If we look at the issue of validity in the most general fashion, the two biggest worries that we have are *confounds* and *artifact*. These two terms are defined in the following way:


- **_Confound_**: A confound is an additional, often unmeasured variable^[The reason why I say that it's unmeasured is that if you *have* measured it, then you can use some fancy statistical tricks to deal with the confound. Because of the existence of these statistical solutions to the problem of confounds, we often refer to a confound that we have measured and dealt with as a *covariate*. Dealing with covariates is a topic for a more advanced course, but I thought I'd mention it in passing, since it's kind of comforting to at least know that this stuff exists.] that turns out to be related to both the predictors and the outcomes. The existence of confounds threatens the internal validity of the study because you can't tell whether the predictor causes the outcome, or if the confounding variable causes it, etc.
- ***Artifact***: A result is said to be "artifactual" if it only holds in the special situation that you happened to test in your study. The possibility that your result is an artifact describes a threat to your external validity, because it raises the possibility that you can't generalise your results to the actual population that you care about.



As a general rule confounds are a bigger concern for non-experimental studies, precisely because they're not proper experiments: by definition, you're leaving lots of things uncontrolled, so there's a lot of scope for confounds working their way into your study. Experimental research tends to be much less vulnerable to confounds: the more control you have over what happens during the study, the more you can prevent confounds from appearing.

However, there's always swings and roundabouts, and when we start thinking about artifacts rather than confounds, the shoe is very firmly on the other foot. For the most part, artifactual results tend to be a concern for experimental studies than for non-experimental studies. To see this, it helps to realise that the reason that a lot of studies are non-experimental is precisely because what the researcher is trying to do is examine human behaviour in a more naturalistic context. By working in a more real-world context, you lose experimental control (making yourself vulnerable to confounds) but because you tend to be studying human psychology "in the wild" you reduce the chances of getting an artifactual result. Or, to put it another way, when you take psychology out of the wild and bring it into the lab (which we usually have to do to gain our experimental control), you always run the risk of accidentally studying something different than you wanted to study: which is more or less the definition of an artifact.

Be warned though: the above is a rough guide only. It's absolutely possible to have confounds in an experiment, and to get artifactual results with non-experimental studies. This can happen for all sorts of reasons, not least of which is researcher error. In practice, it's really hard to think everything through ahead of time, and even very good researchers make mistakes. But other times it's unavoidable, simply because the researcher has ethics (e.g., see \@ref(differentialattrition)). 

Okay. There's a sense in which almost any threat to validity can be characterised as a confound or an artifact: they're pretty vague concepts. So let's have a look at some of the most common examples...


###History effects

***History effects*** refer to the possibility that specific events may occur during the study itself that might influence the outcomes. For instance, something might happen in between a pre-test and a post-test. Or, in between testing participant 23 and participant 24. Alternatively, it might be that you're looking at an older study, which was perfectly valid for its time, but the world has changed enough since then that the conclusions are no longer trustworthy. Examples of things that would count as history effects:


- You're interested in how people think about risk and uncertainty. You started your data collection in December 2010. But finding participants and collecting data takes time, so you're still finding new people in February 2011. Unfortunately for you (and even more unfortunately for others), the Queensland floods occurred in January 2011, causing billions of dollars of damage and killing many people. Not surprisingly, the people tested in February 2011 express quite different beliefs about handling risk than the people tested in December 2010. Which (if any) of these reflects the "true" beliefs of participants? I think the answer is probably both: the Queensland floods genuinely changed the beliefs of the Australian public, though possibly only temporarily. The key thing here is that the "history" of the people tested in February is quite different to people tested in December. 
- You're testing the psychological effects of a new anti-anxiety drug. So what you do is measure anxiety before administering the drug (e.g., by self-report, and taking physiological measures, let's say), then you administer the drug, and then you take the same measures afterwards. In the middle, however, because your labs are in Los Angeles, there's an earthquake, which increases the anxiety of the participants.  


###Maturation effects

As with history effects, ***maturational effects*** are fundamentally about change over time. However, maturation effects aren't in response to specific events. Rather, they relate to how people change on their own over time: we get older, we get tired, we get bored, etc. Some examples of maturation effects:


- When doing developmental psychology research, you need to be aware that children grow up quite rapidly. So, suppose that you want to find out whether some educational trick helps with  vocabulary size among 3 year olds. One thing that you need to be aware of is that the vocabulary size of children that age is growing at an incredible rate (multiple words per day), all on its own. If you design your study without taking this maturational effect into account, then you won't be able to tell if your educational trick works.
- When running a very long experiment in the lab (say, something that goes for 3 hours), it's very likely that people will begin to get bored and tired, and that this maturational effect will cause performance to decline, regardless of anything else going on in the experiment



  
 
###Repeated testing effects

An important type of history effect is the effect of ***repeated testing***. Suppose I want to take two measurements of some psychological construct (e.g., anxiety). One thing I might be worried about is if the first measurement has an effect on the second measurement. In other words, this is a history effect in which the "event" that influences the second measurement is the first measurement itself! This is not at all uncommon. Examples of this include:


- *Learning and practice*: e.g., "intelligence" at time 2 might appear to go up relative to time 1 because participants learned the general rules of how to solve "intelligence-test-style" questions during the first testing session.  
- *Familiarity with the testing situation*: e.g., if people are nervous at time 1, this might make performance go down; after sitting through the first testing situation, they might calm down a lot precisely because they've seen what the testing looks like. 
- *Auxiliary changes caused by testing*: e.g., if a questionnaire assessing mood is boring, then mood at measurement at time 2 is more likely to become "bored", precisely because of the boring measurement made at time 1. 



###Selection bias 

***Selection bias*** is a pretty broad term. Suppose that you're running an experiment with two groups of participants, where each group gets a different "treatment", and you want to see if the different treatments lead to different outcomes. However, suppose that, despite your best efforts, you've ended up with a gender imbalance across groups (say, group A has 80% females and group B has 50% females). It might sound like this could never happen, but trust me, it can. This is an example of a selection bias, in which the people "selected into" the two groups have different characteristics. If any of those characteristics turns out to be relevant (say, your treatment works better on females than males) then you're in a lot of trouble. 

###Differential attrition{#differentialattrition}

One quite subtle danger to be aware of is called ***differential attrition***, which is a kind of selection bias that is caused by the study itself. Suppose that, for the first time ever in the history of psychology, I manage to find the perfectly balanced and representative sample of people. I start running "Dan's incredibly long and tedious experiment" on my perfect sample, but then, because my study is incredibly long and tedious, lots of people start dropping out. I can't stop this: as we'll discuss later in the chapter on research ethics, participants absolutely have the right to stop doing any experiment, any time, for whatever reason they feel like, and as researchers we are morally (and professionally) obliged to remind people that they do have this right. So, suppose that "Dan's incredibly long and tedious experiment" has a very high drop out rate. What do you suppose the odds are that this drop out is random? Answer: zero. Almost certainly, the people who remain are more conscientious, more tolerant of boredom etc than those that leave. To the extent that (say) conscientiousness is relevant to the psychological phenomenon that I care about, this attrition can decrease the validity of my results.

When thinking about the effects of differential attrition, it is sometimes helpful to distinguish between two different types. The first is ***homogeneous attrition***, in which the attrition effect is the same for all groups, treatments or conditions. In the example I gave above, the differential attrition would be homogeneous if (and only if) the easily bored participants are dropping out of all of the conditions in my experiment at about the same rate. In general, the main effect of homogeneous attrition is likely to be that it makes your sample unrepresentative. As such, the biggest worry that you'll have is that the generalisability of the results decreases: in other words, you lose external validity.

The second type of differential attrition is **_heterogeneous attrition_**, in which the attrition effect is different for different groups. This is a much bigger problem: not only do you have to worry about your external validity, you also have to worry about your internal validity too. To see why this is the case, let's consider a very dumb study in which I want to see if insulting people makes them act in a more obedient way. Why anyone would actually want to study that I don't know, but let's suppose I really, deeply cared about this. So, I design my experiment with two conditions. In the "treatment" condition, the experimenter insults the participant and then gives them a questionnaire designed to measure obedience. In the "control" condition, the experimenter engages in a bit of pointless chitchat and then gives them the questionnaire. Leaving aside the questionable scientific merits and dubious ethics of such a study, let's have a think about what might go wrong here. As a general rule, when someone insults me to my face, I tend to get much less co-operative. So, there's a pretty good chance that a lot more people are going to drop out of the treatment condition than the control condition. And this drop out isn't going to be random. The people most likely to drop out would probably be the people who don't care all that much about the importance of obediently sitting through the experiment. Since the most bloody minded and disobedient people all left the treatment group but not the control group, we've introduced a confound: the people who actually took the questionnaire in the treatment group were *already* more likely to be dutiful and obedient than the people in the control group. In short, in this study insulting people doesn't make them more obedient: it makes the more disobedient people leave the experiment! The internal validity of this experiment is completely shot. 



###Non-response bias

**_Non-response bias_** is closely related to selection bias, and to differential attrition. The simplest version of the problem goes like this. You mail out a survey to 1000 people, and only 300 of them reply. The 300 people who replied are almost certainly not a random subsample. People who respond to surveys are systematically different to people who don't. This introduces a problem when trying to generalise from those 300 people who replied, to the population at large; since you now have a very non-random sample. The issue of non-response bias is more general than this, though. Among the (say) 300 people that did respond to the survey, you might find that not everyone answers every question. If (say) 80 people chose not to answer one of your questions, does this introduce problems? As always, the answer is maybe. If the question that wasn't answered was on the last page of the questionnaire, and those 80 surveys were returned with the last page missing, there's a good chance that the missing data isn't a big deal: probably the pages just fell off.  However, if the question that 80 people didn't answer was the most confrontational or invasive personal question in the questionnaire, then almost certainly you've got a problem. In essence, what you're dealing with here is what's called the problem of ***missing data***. If the data that is missing was "lost" randomly, then it's not a big problem. If it's missing systematically, then it can be a big problem.



###Regression to the mean

***Regression to the mean*** is a curious variation on selection bias. It refers to any situation where you select data based on an extreme value on some measure. Because the measure has natural variation, it almost certainly means that when you take a subsequent measurement, that later measurement will be less extreme than the first one, purely by chance. 


Here's an example. Suppose I'm interested in whether a psychology education has an adverse effect on very smart kids. To do this, I find the 20 psych I students with the best high school grades and look at how well they're doing at university. It turns out that they're doing a lot better than average, but they're not topping the class at university, even though they did top their classes at high school. What's going on? The natural first thought is that this must mean that the psychology classes must be having an adverse effect on those students. However, while that might very well be the explanation, it's more likely that what you're seeing is an example of "regression to the mean".  To see how it works, let's take a moment to think about what is required to get the best mark in a class, regardless of whether that class be at high school or at university. When you've got a big class, there are going to be *lots* of very smart people enrolled. To get the best mark you have to be very smart, work very hard, and be a bit lucky. The exam has to ask just the right questions for your idiosyncratic skills, and you have to not make any dumb mistakes (we all do that sometimes) when answering them. And that's the thing: intelligence and hard work are transferrable from one class to the next. Luck isn't. The people who got lucky in high school won't be the same as the people who get lucky at university. That's the very definition of "luck". The consequence of this is that, when you select people at the very extreme values of one measurement (the top 20 students), you're selecting for hard work, skill and luck. But because the luck doesn't transfer to the second measurement (only the skill and work), these people will all be expected to drop a little bit when you measure them a second time (at university). So their scores fall back a little bit, back towards everyone else. This is regression to the mean.

Regression to the mean is surprisingly common. For instance, if two very tall people have kids, their children will tend to be taller than average, but not as tall as the parents. The reverse happens with very short parents: two very short parents will tend to have short children, but nevertheless those kids will tend to be taller than the parents. It can also be extremely subtle. For instance, there have been studies done that suggested that people learn better from negative feedback than from positive feedback. However, the way that people tried to show this was to give people positive reinforcement whenever they did good, and negative reinforcement when they did bad. And what you see is that after the positive reinforcement, people tended to do worse; but after the negative reinforcement they tended to do better. But! Notice that there's a selection bias here: when people do very well, you're selecting for "high" values, and so you should *expect* (because of regression to the mean) that performance on the next trial should be worse, regardless of whether reinforcement is given. Similarly, after a bad trial, people will tend to improve all on their own. The apparent superiority of negative feedback is an artifact caused by regression to the mean [see @Kahneman1973 for discussion].



###Experimenter bias

***Experimenter bias*** can come in multiple forms. The basic idea is that the experimenter, despite the best of intentions, can accidentally end up influencing the results of the experiment by subtly communicating the "right answer" or the "desired behaviour" to the participants. Typically, this occurs because the experimenter has special knowledge that the participant does not -- either the right answer to the questions being asked, or knowledge of the expected pattern of performance for the condition that the participant is in, and so on. The classic example of this happening is the case study of "Clever Hans", which dates back to 1907 [@Pfungst1911; @Hothersall2004]. Clever Hans was a horse that apparently was able to read and count, and perform other human like feats of intelligence. After Clever Hans became famous, psychologists started examining his behaviour more closely. It turned out that -- not surprisingly -- Hans didn't know how to do maths. Rather, Hans was responding to the human observers around him. Because they did know how to count, and the horse had learned to change its behaviour when people changed theirs. 



The general solution to the problem of experimenter bias is to engage in double blind studies, where neither the experimenter nor the participant knows which condition the participant is in, or knows what the desired behaviour is. This provides a very good solution to the problem, but it's important to recognise that it's not quite ideal, and hard to pull off perfectly. For instance, the obvious way that I could try to construct a double blind study is to have one of my Ph.D. students (one who doesn't know anything about the experiment) run the study. That feels like it should be enough. The only person (me) who knows all the details (e.g., correct answers to the questions, assignments of participants to conditions) has no interaction with the participants, and the person who does all the talking to people (the Ph.D. student) doesn't know anything. Except, that last part is very unlikely to be true. In order for the Ph.D. student to run the study effectively, they need to have been briefed by me, the researcher. And, as it happens, the Ph.D. student also knows me, and knows a bit about my general beliefs about people and psychology (e.g., I tend to think humans are much smarter than psychologists give them credit for).  As a result of all this, it's almost impossible for the experimenter to avoid knowing a little bit about what expectations I have. And even a little bit of knowledge can have an effect: suppose the experimenter accidentally conveys the fact that the participants are expected to do well in this task. Well, there's a thing called the "Pygmalion effect": if you expect great things of people, they'll rise to the occasion; but if you expect them to fail, they'll do that too. In other words, the expectations become a self-fulfilling prophesy.


###Demand effects and reactivity

When talking about experimenter bias, the worry is that the experimenter's knowledge or desires for the experiment are communicated to the participants, and that these effect people's behaviour [@Rosenthal1966]. However, even if you manage to stop this from happening, it's almost impossible to stop people from knowing that they're part of a psychological study. And the mere fact of knowing that someone is watching/studying you can have a pretty big effect on behaviour. This is generally referred to as ***reactivity*** or ***demand effects***. The basic idea is captured by the Hawthorne effect: people alter their performance because of the attention that the study focuses on them. The effect takes its name from a the "Hawthorne Works" factory outside of Chicago [see @Adair1984]. A study done in the 1920s looking at the effects of lighting on worker productivity at the factory turned out to be an effect of the fact that the workers knew they were being studied, rather than the lighting.

To get a bit more specific about some of the ways in which the mere fact of being in a study can change how people behave, it helps to think like a social psychologist and look at some of the *roles* that people might adopt during an experiment, but might not adopt if the corresponding events were occurring in the real world:

- The *good participant* tries to be too helpful to the researcher: he or she seeks to figure out the experimenter's hypotheses and confirm them.
- The *negative participant* does the exact opposite of the good participant: he or she seeks to break or destroy the study or the hypothesis in some way.
- The *faithful participant* is unnaturally obedient: he or she seeks to follow instructions perfectly, regardless of what might have happened in a more realistic setting.
- The *apprehensive participant* gets nervous about being tested or studied, so much so that his or her behaviour becomes highly unnatural, or overly socially desirable.


###Placebo effects

The ***placebo effect*** is a specific type of demand effect that we worry a lot about. It refers to the situation where the mere fact of being treated causes an improvement in outcomes. The classic example comes from clinical trials: if you give people a completely chemically inert drug and tell them that it's a cure for a disease, they will tend to get better faster than people who aren't treated at all. In other words, it is people's belief that they are being treated that causes the improved outcomes, not the drug.

###Situation, measurement and subpopulation effects
 
In some respects, these terms are a catch-all term for "all other threats to external validity". They refer to the fact that the choice of subpopulation from which you draw your participants, the location, timing and manner in which you run your study (including who collects the data) and the tools that you use to make your measurements might all be influencing the results. Specifically, the worry is that these things might be influencing the results in such a way that the results won't generalise to a wider array of people, places and measures. 
 
###Fraud, deception and self-deception 

>*It is difficult to get a man to understand something, when his salary depends on his not understanding it.*
>
>-- Upton Sinclair


One final thing that I feel like I should mention. While reading what the textbooks often have to say about assessing the validity of the study, I couldn't help but notice that they seem to  make the assumption that the researcher is honest. I find this hilarious. While the vast majority of scientists are honest, in my experience at least, some are not.^[Some people might argue that if you're not honest then you're not a real scientist. Which does have some truth to it I guess, but that's disingenuous (google the "No true Scotsman" fallacy). The fact is that there are lots of people who are employed ostensibly as scientists, and whose  work has all of the trappings of science, but who are outright fraudulent. Pretending that they don't exist by saying that they're not scientists is just childish.] Not only that, as I mentioned earlier, scientists are not immune to belief bias -- it's easy for a researcher to end up deceiving themselves into believing the wrong thing, and this can lead them to conduct subtly flawed research, and then hide those flaws when they write it up. So you need to consider not only the (probably unlikely) possibility of outright fraud, but also the (probably quite common) possibility that the research is unintentionally "slanted". I opened a few standard textbooks and didn't find much of a discussion of this problem, so here's my own attempt to list a few ways in which these issues can arise are:


- ***Data fabrication***. Sometimes, people just make up the data. This is occasionally done with "good" intentions. For instance, the researcher believes that the fabricated data do reflect the truth, and may actually reflect "slightly cleaned up" versions of actual data. On other occasions, the fraud is deliberate and malicious. Some high-profile examples where data fabrication has been alleged or shown include Cyril Burt (a psychologist who is thought to have fabricated some of his data), Andrew Wakefield (who has been accused of fabricating his data connecting the MMR vaccine to autism) and Hwang Woo-suk (who falsified a lot of his data on stem cell research).  
- ***Hoaxes***. Hoaxes share a lot of similarities with data fabrication, but they differ in the intended purpose. A hoax is often a joke, and many of them are intended to be (eventually) discovered. Often, the point of a hoax is to discredit someone or some field. There's quite a few well known scientific hoaxes that have occurred over the years (e.g., Piltdown man) some of were deliberate attempts to discredit particular fields of research (e.g., the Sokal affair). 
- ***Data misrepresentation***. While fraud gets most of the headlines, it's much more common in my experience to see data being misrepresented. When I say this, I'm not referring to newspapers getting it wrong (which they do, almost always). I'm referring to the fact that often, the data don't actually say what the researchers think they say. My guess is that, almost always, this isn't the result of deliberate dishonesty, it's due to a lack of sophistication in the data analyses. For instance, think back to the example of Simpson's paradox that I discussed in the beginning of these notes. It's very common to see people present "aggregated" data of some kind; and sometimes, when you dig deeper and find the raw data yourself, you find that the aggregated data tell a different story to the disaggregated data. Alternatively, you might find that some aspect of the data is being hidden, because it tells an inconvenient story (e.g., the researcher might choose not to refer to a particular variable). There's a lot of variants on this; many of which are very hard to detect.
- **_Study "misdesign"_**. Okay, this one is subtle. Basically, the issue here is that a researcher designs a study that has built-in flaws, and those flaws are never reported in the paper. The data that are reported are completely real, and are correctly analysed, but they are produced by a study that is actually quite wrongly put together. The researcher really wants to find a particular effect, and so the study is set up in such a way as to make it "easy" to (artifactually) observe that effect. One sneaky way to do this -- in case you're feeling like dabbling in a bit of fraud yourself -- is to design an experiment in which it's  obvious to the participants what they're "supposed" to be doing, and then let reactivity work its magic for you. If you want, you can add all the trappings of double blind experimentation etc. It won't make a difference, since the study materials themselves are subtly telling people what you want them to do. When you write up the results, the fraud won't be obvious to the reader: what's obvious to the participant when they're in the experimental context isn't always obvious to the person reading the paper. Of course, the way I've described this makes it sound like it's always fraud: probably there are cases where this is done deliberately, but in my experience the bigger concern has been with unintentional misdesign. The researcher *believes* ... and so the study just happens to end up with a built in flaw, and that flaw then magically erases itself when the study is written up for publication.
- **_Data mining & post hoc hypothesising_**. Another way in which the authors of a study can more or less lie about what they found is by engaging in what's referred to as "data mining". As we'll discuss later in the class, if you keep trying to analyse your data in lots of different ways, you'll eventually find something that "looks" like a real effect but isn't. This is referred to as "data mining". It used to be quite rare because data analysis used to take weeks, but now that everyone has very powerful statistical software on their computers, it's becoming very common. Data mining per se isn't "wrong", but the more that you do it, the bigger the risk you're taking. The thing that is wrong, and I suspect is very common, is *unacknowledged* data mining. That is, the researcher run every possible analysis known to humanity, finds the one that works, and then pretends that this was the only analysis that they ever conducted. Worse yet, they often "invent" a hypothesis after looking at the data, to cover up the data mining. To be clear: it's not wrong to change your beliefs after looking at the data, and to reanalyse your data using your new "post hoc" hypotheses. What is wrong (and, I suspect, common) is failing to acknowledge that you've done so. If you acknowledge that you did it, then other researchers are able to take your behaviour into account. If you don't, then they can't. And that makes your behaviour deceptive. Bad! 
- **_Publication bias & self-censoring_**. Finally, a pervasive bias is "non-reporting" of negative results. This is almost impossible to prevent. Journals don't publish every article that is submitted to them: they prefer to publish articles that find "something". So, if 20 people run an experiment looking at whether reading *Finnegans Wake* causes insanity in humans, and 19 of them find that it doesn't, which one do you think is going to get published? Obviously, it's the one study that did find that *Finnegans Wake* causes insanity ^[Clearly, the real effect is that only insane people would even try to read *Finnegans Wake.*]. This is an example of a *publication bias*: since no-one ever published the 19 studies that didn't find an effect, a naive reader would never know that they existed. Worse yet, most researchers "internalise" this bias, and end up *self-censoring* their research. Knowing that negative results aren't going to be accepted for publication, they never even try to report them. As a friend of mine says "for every experiment that you get published, you also have 10 failures". And she's right. The catch is, while some (maybe most) of those studies are failures for boring reasons (e.g. you stuffed something up) others might be genuine "null" results that you ought to acknowledge when you write up the "good" experiment. And telling which is which is often hard to do. A good place to start is a paper by @Ioannidis2005 with the depressing title "Why most published research findings are false". I'd also suggest taking a look at work by @Kuhberger2014 presenting statistical evidence that this actually happens in psychology.




There's probably a lot more issues like this to think about, but that'll do to start with. What I really want to point out is the blindingly obvious truth that real world science is conducted by actual humans, and only the most gullible of people automatically assumes that everyone else is honest and impartial. Actual scientists aren't usually *that* naive, but for some reason the world likes to pretend that we are, and the textbooks we usually write seem to reinforce that stereotype.

##Summary

This chapter isn't really meant to provide a comprehensive discussion of psychological research methods: it would require another volume just as long as this one to do justice to the topic. However, in real life statistics and study design are tightly intertwined, so it's very handy to discuss some of the key topics. In this chapter, I've briefly discussed the following topics:


- [Introduction to psychological measurement](#measurement). What does it mean to operationalise a theoretical construct? What does it mean to have variables and take measurements?
- [Scales of measurement and types of variables](#scales). Remember that there are *two* different distinctions here: there's the difference between discrete and continuous data, and there's the difference between the four different scale types (nominal, ordinal, interval and ratio). 
- [Reliability of a measurement](#reliability). If I measure the "same" thing twice, should I expect to see the same result? Only if my measure is reliable. But what does it mean to talk about doing the "same" thing? Well, that's why we have different types of reliability. Make sure you remember what they are.
- [Terminology: predictors and outcomes](#ivdv). What roles do variables play in an analysis? Can you remember the difference between predictors and outcomes? Dependent and independent variables? Etc. 
- [Experimental and non-experimental research designs](#researchdesigns). What makes an experiment an experiment? Is it a nice white lab coat, or does it have something to do with researcher control over variables?
- [Validity and its threats](#validity). Does your study measure what you want it to? How might things go wrong? And is it my imagination, or was that a very long list of possible ways in which things can go wrong? 



All this should make clear to you that study design is a critical part of research methodology. I built this chapter from the classic little book by @Campbell1963, but there are of course a large number of textbooks out there on research design. Spend a few minutes with your favourite search engine and you'll find dozens. 